\chapter*{Abstract}
With Server-Side Swift gaining prominence in the realm of back-end development, the presence of a good JWT library has become extremely important as it enables safe and efficient data transfer between two parties, allowing for the verification of integrity and origin of the data. Vapor's JWTKit package has been the go-to library for developers who need JWT-based authentication in their Swift-based backend. The project behind this thesis aims to improve JWTKit's structure, mainly under the hood.

The thesis begins with an introduction to server-side Swift, explaining how the Apple-made language has evolved from being a closed-source Apple-only-client language to being open-sourced and being used on the server side by several companies. Vapor, although not the only one, is the most popular Swift-based web framework, used by big names, first of them being Apple itself.

The main focus of this thesis is redesigning Vapor's JWT library. This new version brings a number of improvements to the repository, the main one being the removal of Google's BoringSSL C library and its replacement with Swift-only implementations for cryptographic operations. This transition allows the library to reduce external dependencies and potential points of failure, also enhancing maintainability and scalability. Sendable conformance in the package is also explored, which aligns the package with Swift's latest concurrency model and enables thread-safe usage of the library. Finally, the new version also brings an improved API to the package, allowing for custom token de/serialisation, new signature algorithms, and more.

In conclusion, this thesis offers a detailed examination of JWTKit's version 5 release, making it a more valuable asset for developers working on server-side Swift applications.