\chapter*{Conclusion}
\addcontentsline{toc}{chapter}{Conclusion}
This thesis has provided an examination of the Swift programming language, focusing on its use on the server and therefore listing numerous features on why the language is an optimal fit for said environment. In particular Swift's type safety, automatic memory management and modern concurrency model all contribute to it being the perfect fit for the server. Furthermore, the thesis discussed how the presence of the SwiftNIO library promotes the creation of server side frameworks such as Vapor and why this framework is a good choice for creating server side applications. Then, the intricacies of JSON Web Tokens were explored, with an overview of their structure and how their cryptography functions under the hood. Afterwards, the merge of these two technologies in the JWTKit package was discussed and how JWTKit provides great support for using JWTs in Swift based applications. Then, the transition to the last version of the package was evaluated. This transition was long overdue as the presence of C in a Swift library, despite Swift's C interoperability, is obsolete. Therefore, BoringSSL was removed from the library, and along with this the library underwent a number of changes, such as the adoption of Swift 5.5's concurrency model, \swiftinline{Sendable} conformance, new customisation APIs and more. Manually implementing functionalities which were available in BoringSSL but not in SwiftCrypto such as the raw RSA initialisation from modulus, exponent and private exponent was likely the most challenging part. This evolution brought more safety to the package, easier maintainability and general stability. Finally, the new version's performance was evaluated against version 4 and, with not much surprise, performance worsened quite a bit during the transition, likely due to added complexity and abstraction over C constructs. While this does not represent good news for the package, it also offers opportunities for further optimisation and innovation, and thanks to the eradication of C code and the adoption of modern Swift patterns, contribution from the Swift community is promoted and, more than ever, likely.